---
title: "Compare R and Python: basic objects"
output: 
  html_document:
    css: style.css
    toc: true
    toc_depth: 1
---

# Vectors

<div class="left">

## R

`R` comes with a vector class. The definition of a vector is using the `c()` function. 

```{r}
x = c(1, 3, 5)
class(x)
print(x)

y = c("a", "b", "c")
class(y)
print(y)
```
</div>

<div class="right">

## Python

The definition of numeric and character vectors is through `[...]`, a `list` object. 

```{python}
x = [1, 3, 5]
type(x)
print(x)

y = ["a", "b", "c"]
type(y)
print(y)
```

</div>


# Subsetting vectors

The biggest difference between `R` and `Python` vectors is that `R` indexes from 1 and `Python` indexes from 0. 

As a result of this, in `Python`:

+ when subsetting multiple consecutive elements from a list, the left index is inclusive but the right index is exclusive.
+ when subsetting from the beginning of a list, Python doesn't have to specify the first index. `x[0:3]` can be written as `x[:3]`. One can think of this as "getting 3 elements from the vector".

Interestingly, `Python` does not operate on vectors element-wise as you would expect in `R`. To me, this makes `Python` slower to write when one needs to write codes for data analytics.  

<div class="left">

## R

```{r}
x = c(1, 2, 3, 4, 5)
x[1]
x[2]
x[2:4]

x[1:3]
## No equivalent way in R
##
##
##
```

### Extracting multiple elements
```{r}
x[c(1, 3, 5)]
##
##
##
##
##
```


### Filtering using logical
```{r}
x[x > 3]
```

</div>

<div class="right">

## Python

```{python}
x = [1, 2, 3, 4, 5]
x[0]
x[1]
x[1:4]

x[0:3]
x[:3] ## Equivalent to x[0:3]
```

### Extracting multiple elements
```{python}
b = [0, 2, 4]
[x[i] for i in b]
import pandas as pd
list(pd.Series(x)[b])
```

### Filtering using logical

```{python}
print(list(filter(lambda y: y > 3, x)))
```
</div>


<div style="margin-bottom:900px;">
</div>

# Negative indices

+ negative indices carry very different meaning between the two languages. In `R`, the element with the negation will be deleted and in `Python`, the element counting from the end of the vector will be extracted.

+ extracting elements from the end of a vector/list is easier in Python than R because of this behaviour. 

<div class="left">

## R

```{r}
x[-3] ## Deleting the third element 

x[(length(x) - 1):length(x)]
```

</div>

<div class="right">
## Python

```{python}
del x[2]
print(x)

x[-2:] ## Extracting the second last element to the end of the vector
```

</div>

# Consecutive numbers

Again, due to the indexing, Python will require an offset. 

```{r}
1:8
```

```{python}
list(range(1, 9))
```


# Variable assignment

It is very important to note that in Python, the `=` assignment only allows one to make another reference to the same variable. Whereas in R, performing this assignment will overwrite the original object. 

<div class="left">
## R

```{r}
x = 1
y = x
print(y)
y = 2
print(y)
```
</div>

<div class="right">
## Python

```{python}
x = 1
y = x
print(y)
y = 2
print(y)
```
</div>




# Conversion functions

<div class="left">

## R

In `R`, the `class` function shows the class of the object. And almost all functions that converts an object of one class to another is in the form of `as.xyz`.

```{r}
x = 1
print(x)
class(x)

x_chr = as.character(x)
print(x_chr) ## Notice the quotes
class(x_chr)

x_int = as.integer(x_chr)
print(x_int)
class(x_int)
```
</div>

<div class="right">

## Python

In `python`, the `type` function shows the class of the object. The conversion between different classes has a specialised function. 

```{python}
x = 1
print(x)
type(x)

x_chr = str(x)
print(x_chr)
type(x_chr)

x_int = int(x_chr)
print(x_int)
type(x_int)
```
</div>



# List and tuples

<div class="left">

## R

`list` in `R` can very powerful. Each element of a `list` can be literally any object of any class. 

```{r}
x = list(
  numeric = 1,
  character = "cats",
  data_frame = data.frame(x = 1:3, 
                          y = 2:4))

print(x)
class(x)
lapply(x, class)
```

</div>

<div class="right">
## Python

As we have seen above, `list` in `Python` behaves more or less like a vector. 

```{python}
x = [1, "cats"]

print(x)
type(x)
print(list(map(lambda y: type(y), x)))
```


A `tuple` behaves like a `list`, but it is immutable. 

```{python, error = TRUE}
y = tuple(x);

x[0] = 10
print(x)

y[0] = 10
print(y)
```

</div>

# Named vector and dictionary

Suppose we have a matching pair of information in the form of two vectors, and we wish to extract elements of a vector based on another vector. This can of course be done through the usual manipulations on vectors/for loops etc. However, both `R` and `Python` provide different solutions to this seemingly simple task. 

<div class="left">

## R

In `R`, no additional class of object is needed. A `vector` can be given a vector of names, and the subsetting can be done using the names themselves. 

```{r}
x = c("alpha", "bravo", "charlie")
names(x) = c("a", "b", "c")
x[c("a", "c")]
```

</div>

<div class="right">

## Python

In `Python`, a very interesting object class is `dict` (dictionary). A dictionary has two properties, a key and a value. The key behaves much like the `names()` of a vector in `R`. 

```{python}
d = {
"a": "alpha",
"b": "bravo",
"c": "charlie"
}
d["a"]
[d[i] for i in ["a", "c"]]
```

</div>

<div style="margin-bottom:900px;">
</div>

# References

+ https://datacarpentry.org/python-ecology-lesson/03-index-slice-subset/index.html

# Session info

```{r, echo = FALSE}
sessionInfo()
reticulate::py_config()
```
