---
title: "Compare R and Python: basic objects"
output: 
  html_document:
    css: style.css
    toc: true
    toc_depth: 1
---

```{r, echo = FALSE}
reticulate::use_condaenv("r-reticulate")
reticulate::py_config()
```


# Vectors

<div class="left">

## R

`R` comes with a vector class. The definition of a vector is using the `c()` function. 

```{r}
x = c(1, 3, 5)
class(x)
print(x)

y = c("a", "b", "c")
class(y)
print(y)
```
</div>

<div class="right">

## Python

The definition of numeric and character vectors is through `[...]`, a `list` object. 

```{python}
x = [1, 3, 5]
type(x)
print(x)

y = ["a", "b", "c"]
type(y)
print(y)
```

</div>


# Subsetting vectors

The biggest difference between R and Python is that R indexes from 1 and Python indexes from 0. 
As a result of this: 

+ when subsetting multiple consecutive elements from a list, the left index is inclusive but the right index is exclusive.
+ when subsetting from the beginning of a list, Python doesn't have to specify the first index. `x[0:3]` can be written as `x[:3]`.

<div class="left">

## R

```{r}
x = c(1, 2, 3, 4, 5)
x[1]
x[2]
x[2:4]

x[1:3]
```

</div>

<div class="right">
## Python

```{python}
x = [1, 2, 3, 4, 5]
x[0]
x[1]
x[1:4]

x[0:3]
x[:3] ## Equivalent to x[0:3]
```

</div>


<div style="margin-bottom:800px;">
</div>

# Negative indices

+ negative indices carry very different meaning between the two languages. In R, the element with the negation will be deleted and in Python, the reversed element will be extracted.
+ extracting elements from the end of a vector/list is easier in Python than R because of this behaviour. 

<div class="left">

## R

```{r}
x[-3] ## Deleting the third element 

x[(length(x) - 1):length(x)]
```

</div>

<div class="right">
## Python

```{python}
del x[2]
print(x)

x[-2:] ## Extracting the second last element to the end of the vector
```

</div>

# Consecutive numbers

Again, due to the indexing, Python will require an offset. 

```{r}
1:8
```

```{python}
list(range(1, 9))
```


# Variable assignment

It is very important to note that in Python, the `=` assignment only allows one to make another reference to the same variable. Whereas in R, performing this assignment will overwrite the original object. 

<div class="left">
## R

```{r}
x = 1
y = x
print(y)
y = 2
print(y)
```
</div>

<div class="right">
## Python

```{python}
x = 1
y = x
print(y)
y = 2
print(y)
```
</div>

# Conversion functions

<div class="left">

## R

In `R`, the `class` function shows the class of the object. And almost all functions that converts an object of one class to another is in the form of `as.xyz`.

```{r}
x = 1
print(x)
class(x)

x_chr = as.character(x)
print(x_chr) ## Notice the quotes
class(x_chr)

x_int = as.integer(x_chr)
print(x_int)
class(x_int)
```
</div>

<div class="right">

## Python

In `python`, the `type` function shows the class of the object. The conversion between different classes has a specialised function. 

```{python}
x = 1
print(x)
type(x)

x_chr = str(x)
print(x_chr)
type(x_chr)

x_int = int(x_chr)
print(x_int)
type(x_int)
```
</div>


# References

+ https://datacarpentry.org/python-ecology-lesson/03-index-slice-subset/index.html