---
title: "Compare R and Python: data frame manipulations"
output: 
  html_document:
    css: style.css
    toc: true
    toc_depth: 1
---

```{r, echo = FALSE}
library(reticulate)
reticulate::use_virtualenv(virtualenv = "r-reticulate", required = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

---

# Creating a data frame manually

::: {.left}
## R

```{r}
data = data.frame(
  x = 1:5, 
  y = 2:6,
  row.names = c("a", "b", "c", "d", "e"))

print(data)
```
:::

::: {.right}
## Python

```{python}
import pandas as pd
data = pd.DataFrame(
  {"x": list(range(1, 6)),
   "y": list(range(2, 7))}, 
  index = ["a", "b", "c", "d", "e"])
print(data)
```
:::

::: {style="margin-bottom:400px;"}
:::

---

# Reading data frames

::: {.left}
## R

```{r}
library(readr)
data = readr::read_csv("data/penguins_complete.csv")
```
:::

::: {.right}
## Python

```{python}
import pandas as pd
data = pd.read_csv("data/penguins_complete.csv")
```
:::

---

# Saving data frames

::: {.left}
## R

```{r, eval = FALSE}
readr::write_csv(x = data, file = "data/penguins_complete.csv")
```
:::

::: {.right}
## Python

```{python, eval = FALSE}
data.to_csv(path_or_buf = "data/penguins_complete2.csv")
```
:::

## Index-based selection

In `R`, both index-based and label-based subsetting are performed through `[,]`. In `Python`, this is separated into `iloc` and `loc` respectively.

::: {.left}
### R

```{r}
## First row
data[1,]
data[c(1,3,5), c(1,3,5)]
```
:::

::: {.right}
### Python

```{python}
## First row
data.iloc[0,:]
data.iloc[[0,2,4],[0,2,4]]
```
:::

::: {style="margin-bottom:600px;"}
:::

---

## Label-based selection

In `R`, both index-based and label-based subsetting are performed through `[,]`. In `Python`, this is separated into `iloc` and `loc` respectively.

::: {.left}
### R

```{r}
## First five rows from the "species" column
data[1:5, "species"]
```
:::

::: {.right}
### Python

```{python}
## First five rows from the "species" column, return as Series
data.loc[0:4,"species"]
```
:::

::: {style="margin-bottom:400px;"}
:::

## A note about `loc` and `iloc`

In Python, `loc`'s label-querying capabilities means that it will subset **inclusively** on the indices. This means that even if we use the same indices in the code, but depending on which method is used, we can get different results.

```{python}
data.loc[0:4,"species"]
data.iloc[0:4,1]
```

## Conditional selection

::: {.left}
### R

```{r}
## Base R solution
data[(data$species == "Adelie") & (data$bill_depth_mm > 21), ]
##
##
data[data$species %in% c("Adelie", "Chinstrap"), ]
```

```{r}
library(dplyr)
data %>% filter((species == "Adelie") & (bill_depth_mm > 21))
data %>% filter(species %in% c("Adelie", "Chinstrap"))
```
:::

::: {.right}
### Python

```{python}
## Using loc
data.loc[(data.species == "Adelie") & (data.bill_depth_mm > 21)]
##
##
##
substypes = ["Adelie", "Chinstrap"]
data.loc[(data.species.isin(substypes))]
```

```{python}
##
##
## Using query()
data.query('(species == "Adelie") & (bill_depth_mm > 21)')
##
##
##
data.query('species == @substypes')
```
:::

::: {style="margin-bottom:2700px;"}
:::

---

# Adding new columns

::: {.left}
## R

```{r}
data["body_mass_kg"] = data$body_mass_g/1000

## dplyr solution
data %>% mutate(body_mass_kg = body_mass_g/1000)
```
:::

::: {.right}
## Python

```{python}
data["body_mass_kg"] = data.body_mass_g/1000

data.assign(body_mass_kg = lambda x: x.body_mass_g/1000)
```
:::

::: {style="margin-bottom:1000px;"}
:::

---

# Previewing and describing data

::: {.left}
## R

```{r}
head(data)
summary(data)
glimpse(data)
dim(data)

unique(data$species)
##
table(data$species)
```
:::

::: {.right}
## Python

```{python}
data.head()
##
##
##
##
##
##
data.describe()
##
##
##
##
##
##
##
##
##
##
##
##
##
##
##
data.info()
##
##
##
##
##
##
##
data.shape

data.species.unique()
data.species.value_counts()
```
:::

::: {style="margin-bottom:2300px;"}
:::

---

# Grouped operations

The `dplyr` package (a part of the `tidyverse`) contains important functions that makes the manipulation of data.frame very easy. A very important part of the dplyr package is the use of the pipe operator, `%>%`. Together with the use of spaces and line breaks, this often makes the `R` code more readable, so that one can read the code from top to bottom and left to right, i.e. consistent with how we would structure an English sentence. This design framework to make readable codes is a very important reason of why `R` has gained a renewed popularity in data science.

In comparison, `Python` (`pandas`) also adopts similar principle, but the structuring of the codes is not as consistent.

::: {.left}
## R

```{r}
library(dplyr)
data %>% 
  group_by(species) %>% 
  tally()

data %>% 
  group_by(species) %>% 
  summarise(mean_bill_length = mean(bill_length_mm))

data %>% 
  group_by(species) %>% 
  summarise_if(.predicate = is.numeric, .funs = mean)

data %>% 
  group_by(species) %>% 
  summarise(mean_bill_length_top1 = bill_length_mm[1])

data %>% 
  group_by(species, sex) %>% 
  summarise(longest = bill_length_mm[bill_length_mm == max(bill_length_mm)])

data %>% 
  group_by(species) %>% 
  summarise_at(.vars = vars(bill_length_mm),
               .funs = lst(min, max, mean, length))
```
:::

::: {.right}
## Python

```{python}
data.groupby("species").species.count()
##
##
##
data.groupby("species")["bill_length_mm"].mean()
##
##
##
data.groupby("species").mean()
##
##
##
data.groupby("species").apply(lambda x: x.bill_length_mm.iloc[0])
##
##
##
data.groupby(["species", "sex"]).apply(lambda x: x.loc[x.bill_length_mm.idxmax(), "bill_length_mm"])
##
##
##
data.groupby("species").bill_length_mm.agg([min, max, "mean", "size"])
```
:::

::: {style="margin-bottom:2000px;"}
:::

---

# Handling of missing values

::: {.left}
## R

```{r}
library(dplyr)
missing_data = readr::read_csv("data/penguins.csv")
missing_data %>% 
  filter(is.na(bill_length_mm))

missing_data %>% 
  dplyr::mutate(bill_length_mm = coalesce(bill_length_mm, 0))
```
:::

::: {.right}
## Python

```{python}
missing_data = pd.read_csv("data/penguins.csv")
##
##
##
##
##
missing_data[pd.isnull(missing_data.bill_length_mm)]
missing_data.bill_length_mm.fillna(0)
```
:::
