---
title: "Compare R and Python: data frame manipulations"
output: 
  html_document:
    css: style.css
    toc: true
    toc_depth: 1
---

```{r, echo = FALSE}
library(reticulate)
reticulate::use_virtualenv(virtualenv = "r-reticulate", required = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

# Creating a data frame manually 

<div class="left">

## R

```{r}
data = data.frame(
  x = 1:5, 
  y = 2:6,
  row.names = c("a", "b", "c", "d", "e"))

print(data)
```
</div>

<div class="right">

## Python
```{python}
import pandas as pd
data = pd.DataFrame(
  {"x": list(range(1, 6)),
   "y": list(range(2, 7))}, 
  index = ["a", "b", "c", "d", "e"])
print(data)
```
</div>

<div style="margin-bottom:400px;">
</div>

# Reading data frames

<div class="left">

## R

```{r}
library(readr)
data = readr::read_csv("data/penguins_complete.csv")
```
</div>

<div class="right">

## Python
```{python}
import pandas as pd
data = pd.read_csv("data/penguins_complete.csv")
```
</div>

# Saving data frames

<div class="left">

## R

```{r, eval = FALSE}
readr::write_csv(x = data, file = "data/penguins_complete.csv")
```
</div>

<div class="right">

## Python
```{python, eval = FALSE}
data.to_csv(path_or_buf = "data/penguins_complete2.csv")
## 
```
</div>

## Index-based selection
In `R`, both index-based and label-based subsetting are performed through `[,]`. In `Python`, this is separated into `iloc` and `loc` respectively. 

<div class="left">
### R

```{r}
## First row
data[1,]
data[c(1,3,5), c(1,3,5)]
```
</div>

<div class="right">

### Python
```{python}
## First row
data.iloc[0,:]
data.iloc[[0,2,4],[0,2,4]]
```
</div>

<div style="margin-bottom:600px;">
</div>

## Label-based selection
In `R`, both index-based and label-based subsetting are performed through `[,]`. In `Python`, this is separated into `iloc` and `loc` respectively. 

<div class="left">

### R

```{r}
## First five rows from the "species" column
data[1:5, "species"]
```
</div>

<div class="right">

### Python
```{python}
## First five rows from the "species" column, return as Series
data.loc[0:4,"species"]
```
</div>

<div style="margin-bottom:400px;">
</div>

## A note about `loc` and `iloc`

In Python, `loc`'s label-querying capabilities means that it will subset **inclusively** on the indices. This means that even if we use the same indices in the code, but depending on which method is used, we can get different results.  

```{python}
data.loc[0:4,"species"]
data.iloc[0:4,1]
```


## Conditional selection

<div class="left">

### R

```{r}
## Base R solution
data[(data$species == "Adelie") & (data$bill_depth_mm > 21), ]
##
##
data[data$species %in% c("Adelie", "Chinstrap"), ]
```

```{r}
library(dplyr)
data %>% filter((species == "Adelie") & (bill_depth_mm > 21))
data %>% filter(species %in% c("Adelie", "Chinstrap"))
```
</div>


<div class="right">

### Python
```{python}
## Using loc
data.loc[(data.species == "Adelie") & (data.bill_depth_mm > 21)]
##
##
##
substypes = ["Adelie", "Chinstrap"]
data.loc[(data.species.isin(substypes))]
```

```{python}
##
##
## Using query()
data.query('(species == "Adelie") & (bill_depth_mm > 21)')
##
##
##
data.query('species == @substypes')
```
</div>



<div style="margin-bottom:2700px;">
</div>

# Adding new columns

<div class="left">

## R

```{r}
data["body_mass_kg"] = data$body_mass_g/1000

## dplyr solution
data %>% mutate(body_mass_kg = body_mass_g/1000)
```
</div>

<div class="right">

## Python
```{python}
data["body_mass_kg"] = data.body_mass_g/1000

data.assign(body_mass_kg = lambda x: x.body_mass_g/1000)
```
</div>

<div style="margin-bottom:1000px;">
</div>

# Previewing and describing data

<div class="left">

## R

```{r}
head(data)
summary(data)
glimpse(data)
dim(data)

unique(data$species)
##
table(data$species)
```
</div>

<div class="right">

## Python
```{python}
data.head()
##
##
##
##
##
##
data.describe()
##
##
##
##
##
##
##
##
##
##
##
##
##
##
##
data.info()
##
##
##
##
##
##
##
data.shape

data.species.unique()
data.species.value_counts()
```
</div>

<div style="margin-bottom:2000px;">
</div>

# Grouped operations 

The `dplyr` package (a part of the `tidyverse`) contains important functions that makes the manipulation of data.frame very easy. A very important part of the dplyr package is the use of the pipe operator, `%>%`. Together with the use of spaces and line breaks, this often makes the `R` code more readable, so that one can read the code from top to bottom and left to right, i.e. consistent with how we would structure an English sentence. This design framework to make readable codes is a very important reason of why `R` has gained a renewed popularity in data science.

In comparison, `Python` (`pandas`) also adopts similar principle, but the structuring of the codes is not as consistent. 

<div class="left">

## R

```{r}
library(dplyr)
data %>% 
  group_by(species) %>% 
  tally()

data %>% 
  group_by(species) %>% 
  summarise(mean_bill_length = mean(bill_length_mm))

data %>% 
  group_by(species) %>% 
  summarise_if(.predicate = is.numeric, .funs = mean)

data %>% 
  group_by(species) %>% 
  summarise(mean_bill_length_top1 = bill_length_mm[1])

data %>% 
  group_by(species, sex) %>% 
  summarise(longest = bill_length_mm[bill_length_mm == max(bill_length_mm)])

data %>% 
  group_by(species) %>% 
  summarise_at(.vars = vars(bill_length_mm),
               .funs = lst(min, max, mean, length))
```

</div>

<div class="right">

## Python
```{python}
data.groupby("species").species.count()
##
##
##
data.groupby("species")["bill_length_mm"].mean()
##
##
##
data.groupby("species").mean()
##
##
##
data.groupby("species").apply(lambda x: x.bill_length_mm.iloc[0])
##
##
##
data.groupby(["species", "sex"]).apply(lambda x: x.loc[x.bill_length_mm.idxmax(), "bill_length_mm"])
##
##
##
data.groupby("species").bill_length_mm.agg([min, max, "mean", "size"])
```
</div>

<div style="margin-bottom:2000px;">
</div>

# Handling of missing values
<div class="left">

## R

```{r}
library(dplyr)
missing_data = readr::read_csv("data/penguins.csv")
missing_data %>% 
  filter(is.na(bill_length_mm))

missing_data %>% 
  dplyr::mutate(bill_length_mm = coalesce(bill_length_mm, 0))
```

</div>

<div class="right">

## Python
```{python}
missing_data = pd.read_csv("data/penguins.csv")
##
##
##
##
##
missing_data[pd.isnull(missing_data.bill_length_mm)]
missing_data.bill_length_mm.fillna(0)
```
</div>