---
title: "Compare R and Python: data frame manipulations"
output: 
  html_document:
    css: style.css
    toc: true
    toc_depth: 1
---

```{r, echo = FALSE}
library(reticulate)
reticulate::use_virtualenv(virtualenv = "r-reticulate", required = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

# Creating a data frame manually 

<div class="left">

## R

```{r}
data = data.frame(
  x = 1:5, 
  y = 2:6,
  row.names = c("a", "b", "c", "d", "e"))
print(data)
```
</div>

<div class="right">

## Python
```{python}
import pandas as pd
data = pd.DataFrame(
  {"x": list(range(1, 6)),
   "y": list(range(2, 7))}, 
  index = ["a", "b", "c", "d", "e"])
print(data)
```
</div>

<div style="margin-bottom:400px;">
</div>

# Reading data frames

<div class="left">

## R

```{r}
library(readr)
data = readr::read_csv("data/penguins_complete.csv")
```
</div>

<div class="right">

## Python
```{python}
import pandas as pd
data = pd.read_csv("data/penguins_complete.csv")
```
</div>

# Saving data frames

<div class="left">

## R

```{r, eval = FALSE}
readr::write_csv(x = data, file = "data/penguins_complete.csv")
```
</div>

<div class="right">

## Python
```{python, eval = FALSE}
data.to_csv(path_or_buf = "data/penguins_complete2.csv")
```
</div>

# Indexing and subsetting

<!-- ## Column-wise subsetting -->

<!-- <div class="left"> -->

<!-- ### R -->

<!-- ```{r} -->
<!-- ## A smaller dataframe/tibble -->
<!-- data[,"species"] -->
<!-- ## A vector -->
<!-- head(data$species)  -->
<!-- ``` -->
<!-- </div> -->

<!-- <div class="right"> -->

<!-- ### Python -->
<!-- ```{python} -->
<!-- ## Both are pandas Series -->
<!-- data["species"] -->
<!-- data.species -->

<!-- ``` -->
<!-- </div> -->


## Index-based selection

<div class="left">

### R

```{r}
## First row
data[1,]
## Arbitary rows/columns through indexing
data[c(1,3,5), c(1,3,5)]
```
</div>

<div class="right">

### Python
```{python}
## First row
data.iloc[0,:]
## Arbitary rows/columns through indexing
data.iloc[[0,2,4],[0,2,4]]
```
</div>


## Label-based selection

<div class="left">

### R

```{r}
## First five rows from the "species" column
data[1:5, "species"]
```
</div>

<div class="right">

### Python
```{python}
## First five rows from the "species" column, return as Series
data.loc[0:4,"species"]
```
</div>

## A note about `loc` and `iloc`

`loc`'s label-querying capabilities means that it will subset **inclusively** on the indices. 
```{python}
data.loc[0:4,"species"]
data.iloc[0:4,1]
```


## Conditional selection

<div class="left">

### R

```{r}
## Base R solution
data[(data$species == "Adelie") & (data$bill_depth_mm > 21), ]
data[data$species %in% c("Adelie", "Chinstrap"), ]

library(dplyr)
data %>% filter((species == "Adelie") & (bill_depth_mm > 21))
data %>% filter(species %in% c("Adelie", "Chinstrap"))
```
</div>

<div class="right">

### Python
```{python}
## Using loc
data.loc[(data.species == "Adelie") & (data.bill_depth_mm > 21)]
data.loc[(data.species.isin(["Adelie", "Chinstrap"]))]

## Using query()
data.query('(species == "Adelie") & (bill_depth_mm > 21)')
```
</div>


# Adding new columns

<div class="left">

## R

```{r}
data["body_mass_kg"] = data$body_mass_g/1000

## dplyr solution
data %>% mutate(body_mass_kg = body_mass_g/1000)
```
</div>

<div class="right">

## Python
```{python}
data["body_mass_kg"] = data.body_mass_g/1000

data.assign(body_mass_kg = lambda x: x.body_mass_g/1000)
```
</div>

# Reading data frames (chunked)

# Previewing and describing data

<div class="left">

## R

```{r}
head(data)
summary(data)
glimpse(data)
dim(data)

unique(data$species)
table(data$species)
```
</div>

<div class="right">

## Python
```{python}
data.head()
data.describe()
data.info()
data.shape

data.species.unique()
data.species.value_counts()
```
</div>


# Grouped operations 


<div class="left">

## R

```{r}
library(dplyr)
data %>% 
  group_by(species) %>% 
  tally()

data %>% 
  group_by(species) %>% 
  summarise(mean_bill_length = mean(bill_length_mm))

data %>% 
  group_by(species) %>% 
  summarise_if(.predicate = is.numeric, .funs = mean)

data %>% 
  group_by(species) %>% 
  summarise(mean_bill_length_top1 = bill_length_mm[1])

data %>% 
  group_by(species, sex) %>% 
  summarise(longest = bill_length_mm[bill_length_mm == max(bill_length_mm)])

data %>% 
  group_by(species) %>% 
  summarise_at(.vars = vars(bill_length_mm),
               .funs = lst(min, max, mean, length))
```

</div>

<div class="right">

## Python
```{python}
data.groupby("species").species.count()
data.groupby("species")["bill_length_mm"].mean()
data.groupby("species").mean()
data.groupby("species").apply(lambda x: x.bill_length_mm.iloc[0])
data.groupby(["species", "sex"]).apply(lambda x: x.loc[x.bill_length_mm.idxmax(), "bill_length_mm"])
data.groupby("species").bill_length_mm.agg([min, max, "mean", "size"])
```
</div>

# Handling of missing values



<div class="left">

## R

```{r}
library(dplyr)
missing_data = readr::read_csv("data/penguins.csv")
missing_data %>% 
  filter(is.na(bill_length_mm))

missing_data %>% 
  dplyr::mutate(bill_length_mm = coalesce(bill_length_mm, 0))
```

</div>

<div class="right">

## Python
```{python}
missing_data = pd.read_csv("data/penguins.csv")
missing_data[pd.isnull(missing_data.bill_length_mm)]
missing_data.bill_length_mm.fillna(0)
```
</div>