---
title: "Compare R and Python: functions and methods"
output: 
  html_document:
    css: style.css
    toc: true
    toc_depth: 1
---

```{r, echo = FALSE}
library(reticulate)
reticulate::use_virtualenv(virtualenv = "r-reticulate", required = TRUE)
```

# Self-defined Functions

<div class="left">

## R

In `R`, a function can be defined using `function` and braces are necessary to enclose the function body. 

```{r}
square = function(x){
  return(x ** 2)
}

square(10)
```
</div>

<div class="right">

## Python

In `python`, a function can be defined using `def`, and the body is recognised through the indentations. Four spaces is commonly used for the indentation. 

```{python}
def square(x):
    new_value = x ** 2
    return new_value
    
square(10)
```

</div>

# List comprehensions

Generation of special lists/vectors is a common task in any programming language. 

<div class="left">

## R

```{r}
x = 1:20
(2*x)[x %% 2 == 0]
ifelse(x %% 2 == 0, 2*x, 0)
```
</div>

<div class="right">

## Python

In `python`, list comprehensions is a powerful way to generate a list in the coding style of a for loop. 

```{python}
[2*x for x in range(1, 21) if x % 2 == 0]
[2*x if x % 2 == 0 else 0 for x in range(1, 21)]
```

</div>

# Error handling


# Mapping

<div class="left">

## R

```{r}
x = 1:10
purrr::map_dbl(x, ~.x^2)
```
</div>

<div class="right">

## Python

In `python`, list comprehensions is a powerful way to generate a list in the coding style of a for loop. 

```{python}
x = list(range(1, 11))
list(map(lambda num: num**2, x))
```

</div>
