---
title: "Compare R and Python: model tuning"
output: 
  html_document:
    css: style.css
    toc: true
    toc_depth: 1
---

For this comparison, I will compare the `scikit-learn` library in `Python` and the newly developed `tidymodels` meta-package in `R`. The main reason that I have chosen these two is because they share a lot of similarities and imposed strict frameworks in data pre-processing, modelling and evaluations. 

The data that I will use is the `penguins` data from the `R` package `palmerpenguins`, which you can learn more about [here](https://allisonhorst.github.io/palmerpenguins/articles/intro.html). The response variable is a `factor` variable, `species`, indicating the species of a penguin. The other predictor variables are a mix of both numeric and factor variables. For convenience, I have reduced the number of species to two and extracted the data below in a CSV format so that Python can also use this data through `pd.read_csv`.

```{r, eval = FALSE}
library(palmerpenguins)
library(tidyverse)

penguins %>% 
  na.omit %>% 
  dplyr::filter(species %in% c("Adelie", "Chinstrap")) %>% 
  readr::write_csv(path = "data/penguins_complete.csv")
```

# Importing data and getting a summary 

<div class="left">

## `R`: `tidyverse`

```{r, message = FALSE}
library(tidyverse)
penguins = readr::read_csv(file = "data/penguins_complete.csv")
penguins %>% colnames()
```

</div>

<div class="right">

## `Python`: `pandas`

```{python}
import pandas as pd
penguins = pd.read_csv("data/penguins_complete.csv")
penguins.columns

feature_set = ['bill_length_mm', 'bill_depth_mm', 'flipper_length_mm', 'body_mass_g']
X = penguins[feature_set]
y = penguins.species
```

</div>

<div style="margin-bottom:1000px;">
</div>

# Grid search

## R

## Python

```{python}
from sklearn.tree import DecisionTreeClassifier
from sklearn.model_selection import GridSearchCV
import numpy as np

param_grid = {'max_depth': np.arange(1, 10)} 
dtc_model = DecisionTreeClassifier(random_state = 0)
dtc_cv = GridSearchCV(dtc_model, param_grid, cv = 5) 
dtc_cv.fit(X, y)
dtc_cv.best_params_
```

