---
title: "Compare R and Python: workflows"
output: 
  html_document:
    css: style.css
    toc: true
    toc_depth: 1
---

```{r, echo = FALSE}
library(reticulate)
reticulate::use_virtualenv(virtualenv = "r-reticulate", required = TRUE)
```

For this comparison, I will compare the `scikit-learn` library in `Python` and the newly developed `tidymodels` meta-package in `R`. The main reason that I have chosen these two is because they share a lot of similarities and imposed strict frameworks in data pre-processing, modelling and evaluations. 

The data that I will use is the `penguins` data from the `R` package `palmerpenguins`, which you can learn more about [here](https://allisonhorst.github.io/palmerpenguins/articles/intro.html). The response variable is a factor variable, `species`, indicating the species of a penguin. The other predictor variables are a mix of both numeric and factor variables. For convenience, I have extracted the data below in a CSV format so that Python can also use this data through `pd.read_csv`

```{r, eval = FALSE}
library(palmerpenguins)
library(tidyverse)
readr::write_csv(penguins %>% na.omit, path = "data/penguins_complete.csv")
```

```{r, echo = FALSE}
reticulate::use_condaenv("r-reticulate")
reticulate::py_config()
reticulate::py_module_available("pandas")
reticulate::py_module_available("scikit-learn")
```



# Importing data and getting a summary 

<div class="left">

## `R`: `tidyverse`

It is hard for me to remember how to use `R` without `tidyverse`. And by `tidyverse`, I am referring to both the package and the whole framework that it has established and keeps proliferating through many community-built `R` packages. 

Here, we will use the `readr` and `dplyr` packages to import the data and quickly get a summary of the data. 

```{r, message = FALSE}
library(tidyverse)
penguins = readr::read_csv(file = "data/penguins_complete.csv")
penguins %>% glimpse()
```

</div>

<div class="right">

## `Python`: `pandas`

```{python}
import pandas as pd
penguins = pd.read_csv("data/penguins_complete.csv")
penguins.info()
penguins.describe()
```

</div>

<div style="margin-bottom:1000px;">
</div>

# Splitting data into train-test set

<div class="left">

## `R`: `rsample`

```{r, message = FALSE}
library(tidymodels)
set.seed(2020)
penguins = penguins %>% 
  dplyr::mutate(across(.cols = is.character, .fns = as.factor))

penguins_split = rsample::initial_split(data = penguins, prop = 0.1)
penguins_split

penguins_train = penguins_split %>% training()
penguins_test = penguins_split %>% testing()
```

</div>

<div class="right">
## `Python`: `train_test_split` from `sklearn.model_selection`

```{python}
from sklearn.model_selection import train_test_split
from pandas import get_dummies

y = penguins["species"].values
penguins_x = pd.get_dummies(penguins.drop("species", axis = 1))
x = penguins_x.values

x_train, x_test, y_train, y_test = train_test_split(x, y, train_size = 0.1, random_state = 123)
```

</div>

<div style="margin-bottom:800px;">
</div>

# Setting up a recipe

<div class="left">
## `R`: `recipes`
```{r}
penguins_recipe = recipes::recipe(species ~ ., data = penguins_train) %>% 
  step_knnimpute(all_predictors()) %>% 
  prep()

penguins_recipe
```

</div>

<div class="right">
## `Python`: None
</div>

<div style="margin-bottom:800px;">
</div>

# Defining a (random forest) model

<div class="left">

## `R`: `parsnip`

```{r}
rf_model = rand_forest() %>%
  set_args(trees = 100) %>% 
  set_engine("randomForest") %>% 
  set_mode("classification")
```

</div>

<div class="right">

## `Python`: `sklearn.ensemble`

```{python}
from sklearn.ensemble import RandomForestClassifier

steps = [('rf', RandomForestClassifier(n_estimators = 100))]
```

</div>

<div style="margin-bottom:800px;">
</div>

# Setting up a workflow/pipeline

<div class="left">
## `R`: `workflowr`

```{r}
rf_workflow = workflow() %>% 
  add_recipe(penguins_recipe) %>% 
  add_model(rf_model)

rf_pred_tbl = predict(rf_workflow %>% fit(data = penguins_train), 
                      new_data = penguins_test)

rf_pred_tbl = rf_pred_tbl %>% 
  dplyr::mutate(true = penguins_test$species %>% factor)

rf_pred_tbl
```

</div>

<div class="right">

## `Python`: `Pipeline` from `sklearn.pipeline`

```{python}
from sklearn.pipeline import Pipeline
pipeline = Pipeline(steps = steps)

rf_pipeline_model = pipeline.fit(x_train, y_train)

rf_pred_tbl = pd.DataFrame(
{"true": y_test,
"pred": rf_pipeline_model.predict(x_test)})

rf_pred_tbl
```

</div>

<div style="margin-bottom:800px;">
</div>

# Evaluation

<div class="left">
## `R`: `yardstick`
```{r}
multi_metric = metric_set(accuracy, precision, recall, f_meas)
multi_metric(data = rf_pred_tbl, truth = true, estimate = ".pred_class")
```
</div>

<div class="right">
## `Python`: `sklearn.metrics`

```{python}
from sklearn.metrics import classification_report

cr = classification_report(
y_true = rf_pred_tbl.true, 
y_pred = rf_pred_tbl.pred)

print(cr)
```
</div>